\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\ or \\ Charting antigenic drift and evolutionary success of seasonal influenza lineages

### 2013-01-15 flux.tex
Move data details to supplement.

\subsection*{Posterior samples}

@@

\subsection*{Genetic, antigenic and surveillance data}

We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
However, the majority of isolates were present from 2002 to 2007. 
Because we are interested in longer-term antigenic evolution, we subsampled the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
This censoring left 429 virus isolates, 519 serum isolates and 10,097 HI measurements.

Antigenic data for influenza A/H1N1 was collected from previous publications \cite{Kendal78,Webster79,Nakajima79,Nakajima81,Chakraverty82,Pereira82,Chakraverty86,Cox83,Daniels85,Raymond86,Stevens87,Donatelli93,Hay01,Daum02,McDonald07,Barr10} and NIMR vaccine strain selection reports for 2002--2010 \cite{NIMR02,NIMR03,NIMR04,NIMRFeb05,NIMRSep05,NIMRMarch06,NIMRSep06,NIMRMarch07,NIMRSep07,NIMRMarch08,NIMRSep08,NIMRFeb09,NIMRFeb10}.
The same procedure was followed as was followed for H3N2 to match sequence data and to subsample antigenic comparisons.
This procedure yielded 124 virus isolates, 77 serum isolates and 1882 HI measurements over the course of 1977 to 2009.

Antigenic comparisons for influenza B/Victoria were collated from previous publications \cite{Rota90, Hay01, Muyanga01, Shaw02, Ansaldi04, Puzelli04, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
Here, the sequence matching and subsampling procedure yielded 179 virus isolates, 70 serum isolates and 2003 HI measurements over the course of 1986 to 2011.

Antigenic comparisons for influenza B/Yamagata were collected from previous publications \cite{Rota90, Kanegae90, Nakajima92, Nerome98, Hay01, Muyanga01, Nakagawa02, Abed03, Ansaldi03, Ansaldi04, Matsuzaki04, Puzelli04, Shaw02, Xu04, Barr06, Daum06, Lin07} and vaccine strain selection reports for 2002--2012 \cite{AusWHO06, NIMR02, NIMR03, NIMR04, NIMRFeb05, NIMRSep05, NIMRMarch06, NIMRSep06, NIMRMarch07, NIMRSep07, NIMRMarch08, NIMRFeb09, NIMRSep09, NIMRFeb10, NIMRSep10, NIMRFeb11, NIMRSep11, NIMRFeb12}.
For B/Yamagata, the matching and subsampling procedure resulted in 174 virus isolates, 69 serum isolates and 1962 HI measurements over the course of 1987 to 2011.

Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
As an example, one report states ``collaborating laboratories in the United States tested 195,744 respiratory specimens for influenza viruses, 27,682 (14\%) of which were positive. Of these, 18,175 (66\%) were positive for influenza A viruses, and 9,507 (34\%) were positive for influenza B viruses. Of the 18,175 specimens positive for influenza A viruses, 7,631 (42\%) were subtyped; 6,762 (87\%) of these were seasonal influenza A (H1N1) viruses, and 869 (13\%) were influenza A (H3N2) viruses.''
In this case, we estimate the relative proportion of A/H3N2 of the four clades as $0.66 \times 0.13 = 0.09$.
Similar calculations were performed for A/H1N1, B/Vic and B/Yam.

\subsection*{Implementation considerations}

There were a couple of issues we encountered in applying the BMDS model to HI data from A/H3N2, A/H1N1, B/Vic and B/Yam.
The first issue lay in summarizing the posterior distribution of virus and serum locations.
Although our MCMC procedure samples from the posterior, sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC virus locations may shift.
Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
However, local isometries are often still a problem.
For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
Here, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.

>>>>

\subsection*{Posterior samples and summaries}

@@

There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
However, local isometries are often still a problem.
For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.

\subsection*{Data and implementation}

We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications and WHO vaccine strain selection reports (see Supp.\ Text).
Hemagglutinin sequences were compiled from the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID}.
Antigenic and genetic data were subsampled to give similar quantities of data across years of the analysis.
Data gathering procedures resulted in 402 virus strains A/H3N2, 115 virus strains for A/H1N1, 179 virus strains for B/Vic and 174 virus strains for B/Yam.
Surveillance data was obtained from the Centers of Disease Control and Prevention FluView Influenza Reports from the yearly summaries of influenza seasons 1997--1998 to 2010--2011 \cite{CDCReports}.
